---
title: "Assignment 2: Explore Gapminder and use dplyr"
output: html_document
---
# Exercise 1: Basic `dplyr`

```{r load, warning=FALSE, echo=FALSE}
library(gapminder)
library(tidyverse)
```


## 1.1
Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r 1.1, warning=FALSE, echo=FALSE}
filtered <- gapminder %>% 
  filter(year>=1970 & year<=1980) %>% 
  filter(country=='Norway' | country=='Sweden' | country=='Denmark')
filtered
```

## 1.2
Use the pipe operator %>% to select “country” and “gdpPercap” from your filtered dataset in 1.1.
```{r 1.2, warning=FALSE, echo=FALSE}
filtered %>% 
  select(country, gdpPercap)
```

## 1.3
Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the lag() or diff() functions useful.

<!--
add extra column that's the difference
then select columns where the difference is negative
-->
```{r 1.3, warning=FALSE, echo=FALSE}
gapminder %>% 
  group_by(country) %>% 
  arrange(year) %>% 
  mutate(change=lifeExp-lag(lifeExp, 1)) %>% 
  filter(change<0)
```

## 1.4
Exercise chosen:

Filter gapminder so that it shows the max GDP per capita experienced by each country. Hint: you might find the max() function useful here.

```{r 1.4, warning=FALSE, echo=FALSE}
gapminder %>% 
  group_by(country) %>% 
  filter(gdpPercap==max(gdpPercap))
```

## 1.5
Produce a scatterplot of Canada’s life expectancy vs. GDP per capita using ggplot2, without defining a new variable. That is, after filtering the gapminder data set, pipe it directly into the ggplot() function. Ensure GDP per capita is on a log scale.

```{r 1.5, warning=FALSE, echo=FALSE}
gapminder %>% 
  filter(country=='Canada') %>% 
  ggplot(aes(x=gdpPercap, y=lifeExp)) + geom_point() + scale_x_log10()
```


# Exercise 2: Explore individual variables with `dplyr`

Pick one categorical variable and one quantitative variable to explore. Answer the following questions in whichever way you think is appropriate, using dplyr:

Variables chosen: continent (qualitative), pop (quantitative)


What are possible values (or range, whichever is appropriate) of each variable?
```{r 2A, warning=FALSE}
#Possible values for country:
gapminder %>% 
  distinct(continent)

#Range of pop:
gapminder %>% 
  select(pop) %>% 
  range() %>% 
```

What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

```{r 2B, warning=FALSE}
cont_count <- gapminder %>% 
  select(continent) %>% 
  group_by(continent) %>% 
  count()
ggplot(cont_count, aes(x=continent, y=n))+geom_col()
pie(cont_count$n, labels=cont_count$continent)

hist(gapminder$pop, breaks=50)
```

Feel free to use summary stats, tables, figures.

