---
title: 'Assignment 03: dplyr/ggplot2 Part II'
output: html_document
---

```{r}
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
```


Pick three of the six tasks below, and produce:

    a tibble, using dplyr as your data manipulation tool;
    an accompanying plot of data from the tibble, using ggplot2 as your visualization tool; and
    some dialogue about what your tables/figures show (doesn’t have to be much).

Task Option 1

Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.


## Task Option 2

*Get the maximum and minimum of GDP per capita for all continents.*

This question seems a bit ambiguous. Here are some different interpretations I came up with, and the solutions to each:

- Maximum and minimum gdpPercap values in Gapminder across all continents
  - So here, we would only be looking for two numbers... the maximum and minimum gdpPercap entries in the whole dataframe.
  
```{r}
gapminder %>% 
  summarize(max_gdp = max(gdpPercap), min_gdp = min(gdpPercap)) %>% 
  kable()
```
  
- Or is it the maximum and minimum GDP entries for each continent? 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(max_gdp = max(gdpPercap), min_gdp = min(gdpPercap)) %>% 
  kable()
```

- This version shows the Gapminder entries that contained the maximum or minimum GDP per capita **of all time** (at least within the limits of the dataframe) for that continent, including the country and year that it occurred.

```{r}
per_cont <- gapminder %>% 
  group_by(continent) %>% 
  filter(gdpPercap==min(gdpPercap) | gdpPercap==max(gdpPercap)) %>% 
  mutate(type=if_else(gdpPercap==max(gdpPercap), paste("max"), paste("min")))
kable(per_cont)
```

  - Here's a bar graph edition. I tried using a logarithmic y axis scale... didn't like it. The max and min bars became too close together and made it difficult to make a proper comparison.
  
```{r}
per_cont %>% 
  group_by(continent, type) %>% 
  ggplot(aes(x=continent, y=gdpPercap))+geom_bar(stat="identity", aes(fill=type), position=position_dodge2())
```

- We can also take the task request to mean the most recent maximum and minimum GDP per capita for each continent. This way, we're eliminating time as a confounding factor.

```{r}
per_cont_year <- gapminder %>% 
  group_by(continent, year) %>% 
  filter(gdpPercap==min(gdpPercap) | gdpPercap==max(gdpPercap)) %>% 
  mutate(type=if_else(gdpPercap==max(gdpPercap), paste("max"), paste("min")))
datatable(per_cont_year)
per_cont_year %>% 
  group_by(continent, type) %>% 
  arrange(year) %>% 
  ggplot(aes(x=continent, y=gdpPercap))+geom_bar(stat="identity", aes(fill=type), position=position_dodge2())
```

  - This grouped bar graph tries to cram the data from all 12 of the years included in the dataframe into one graph by breaking up each solid bar, first into min GDP and max GDP types, then into several more tiny bars to represent each year. It's a little wacky, but it gets the point across. 
  - We can infer that Asia has experienced massive fluctuations in its maximum GDP per capita, while maximum GDP per capita has been steadily on the rise in Europe, the Americas, and Oceania.
  - One more time, as a line plot!
```{r}
per_cont_year %>% 
  group_by(year, continent) %>% 
  filter(type=="min") %>% 
  ggplot(aes(x=year, y=gdpPercap, color=continent))+geom_line()
per_cont_year %>% 
  group_by(year, continent) %>% 
  filter(type=="max") %>% 
  ggplot(aes(x=year, y=gdpPercap, color=continent))+geom_line()
```
  


Task Option 3

Look at the spread of GDP per capita within the continents.


Task Option 4

Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.


Task Option 5

How is life expectancy changing over time on different continents?


Task Option 6

Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.
