---
title: 'Exercise 3: R Markdown Presentation Slides'
author: "Tracy Wang"
date: "September 2019"
output:
  beamer_presentation:
    incremental: no
autosize: yes
---

# Slides!

This will be a presentation version of the Orange dataset exploration presented in Exercise 2.

# Orange

Orange is a dataset about the growth of five orange trees. This is a record of their trunk circumferences (in cm) at certain ages (in days).

The headers for the Orange dataset, as determined through the names() function, are:

```{r load, warning=FALSE, echo=FALSE}
library(tibble)
library(gapminder)
library(DT)

names(Orange)
```

# Summary

Here is a summary of the Orange dataset using the summary() function:

```{r summary, warning=FALSE, echo=FALSE}
summary(Orange)
```

# Pie

In the Orange dataset, each individual tree is labeled with an identifying number, ranging from 1 to 5. The following pie chart represents the relative number of times each tree is represented in the dataset:

```{r pie, warning=FALSE, echo=FALSE}
pie(table(Orange$Tree), edges=200, radius=0.8)

```

Each tree has an equal number of data points.


# Plot

And here is a plot of tree circumference versus age:

```{r plot, warning=FALSE, echo=FALSE}
plot(x=Orange[,2], y=Orange[,3], xlab="Age (days)", ylab="Circumference (cm)", main="Tree size with age", type='p')
abline(lm(circumference~age, data=Orange), col='red')
```

# Observations

- Obviously we cannot conclude this is a linear relationship. 
- However, the goal here was simply to add a trendline, and that was a success!


# The end

